Well-quasi-orderings on word languages

### 2025-10-14 src/infixes.tex
new: maybe working proof

\begin{lemma}
  \label{morphic-uur:lem}
    Given a \kl{pure morphic sequence} $w \in \Sigma^{\Nat}$, one can decide
    whether it is \kl{ultimately uniformly recurrent}.
\end{lemma}
\begin{proof}
  On the one hand, one can guess a bound $N_0$ and check that 
  the suffix of the word $w_{\geq N_0}$ is \kl{uniformly recurrent},
  since the suffix of a \kl{morphic sequence} is also a \kl{morphic sequence},
  and \kl{uniform recurrence} is decidable for \kl{morphic sequences}
  \cite[Theorem 2]{DURAND13}.

  Let us now prove that a specific choice of $N_0$ is sufficient. Let us write
  $w \defined f^\omega(a)$ where $f \colon \Sigma^* \to \Sigma^*$ is a
  \kl(words){morphism} and $a \in \Sigma$ is a letter such that $f(a)$ starts
  with $a$. Let us also write $f(a) = a s$, where $s \in \Sigma^+$ is a
  non-empty word. Notice that $w = a s f(s) f^2(s) \cdots$. Without loss of
  generality, one can assume that $|f^n(s)| \to \infty$ when $n \to \infty$,
  otherwise $w$ is ultimately periodic, and therefore \kl{ultimately uniformly
  recurrent}. Since one can algorithmically detect if this is the case, we
  assume that it is not.

  Because $\Sigma$ is finite, there exists a finite set of letters $A \subseteq
  \Sigma$ and a suffix $w'$ of $w$ such that $w' \in A^*$ and every letter of
  $w'$ appears infinitely often in $w'$. Furthermore, one can compute this set
  $A$ and a bound $N_0$ such that $f^n(s) \in A^*$ for every $n \geq N_0$. We
  claim that $w$ is \kl{ultimately uniformly recurrent} if and only if
  $f^{N_0}(s) f^{N_0 + 1}(s) \cdots$ is \kl{uniformly recurrent}.
  One direction being obvious, we assume that $w$ is \kl{ultimately uniformly
  recurrent}, and let us prove that $f^{N_0}(s) f^{N_0 + 1}(s) \cdots$ is
  \kl{uniformly recurrent}.

  Let us assume that $w'' \defined f^{N_1}(s) f^{N_1 + 1}(s) \cdots$ where $N_1
  \geq N_0$ is \kl{uniformly recurrent}. Let us show that $f^{N_0}(s)$ is an
  \kl{infix} of $w''$ if and only if there exists a letter $b$ in $s$ such that
  $f^{N_0}(s) \infleq f^{k} (b)$ for some finite number $k$.

  Assume that $f^{N_0}(s)$ is an \kl{infix} of $w''$. Recall that $|f^n(s)| \to
  \infty$ when $n \to \infty$, hence, there must exist a letter $b$ in $s$ such
  that $|f^n(b)| \to \infty$ when $n \to \infty$. Because every letter of
  $f^{N_0}(s)$ appears infinitely often in $w''$, $f^n(b)$ is a growing
  sequence of infixes of $w''$. Since $w''$ is \kl{uniformly recurrent}, there
  exists a bound $K$ such that $f^{N_0}(s)$ is an \kl{infix} of every $f^k(b)$
  for $k \geq K$. Conversely, if there exists a letter $b$ in $s$ such that
  $f^{N_0}(s) \infleq f^{k} (b)$ for some finite number $k$, then $f^{N_0}(s)$
  is an \kl{infix} of $w''$ because $f^n(b)$ is an infix of $w''$ for every $n
  \in \Nat$.

  Because $w''$ is \kl{uniformly recurrent}, there exists a bound $K$ such that
  $f^{N_1}(s) \infleq f^k(b)$ for every $k \geq K$.
  \todo[inline]{Cannot finish the proof...}

>>>>

\begin{lemma}
  \label{pure-morphic-uur:lem}
  Let $w \in \Sigma^{\Nat}$ be a \kl{pure morphic sequence}
  generated by a morphism $f \colon \Sigma^* \to \Sigma^*$ and a letter $a \in
  \Sigma$ such that $f(n)$ starts with $a$.
  Assume furthermore that $f$ is \kl(morphic){growing}, i.e., that 
  $f(a) = a s$ where $s \in \Sigma^+$ is a non-empty word
  such that $|f^n(s)| \to \infty$ when $n \to \infty$.
  Then, the following are equivalent:
  \begin{itemize}
    \item $w$ is \kl{ultimately uniformly recurrent},
    \item \begin{enumerate}
          \item 
            for every letters $b,c \in \Sigma$,
            if $b$ is \kl(morphic){growing} (i.e., $|f^n(b)| \to \infty$ when $n \to \infty$)
            and both $b$ and $c$ appear infinitely often in $w$, 
            then $c$ appears in $f^m(b)$ for some $m \in \Nat$,
          \item
            there are only finitely many 
            \kl(worphic){bounded words}
            (i.e., words $u \in \Sigma^*$ such that $|f^n(u)|$ is bounded when $n \to \infty$)
            that appear as infixes of $w$.
    \end{enumerate}
  \end{itemize}
\end{lemma}
\begin{proof}
  Assume that $w$ is \kl{ultimately uniformly recurrent}.
  We have $N_0$ such that $w_{\geq N_0}$ is \kl{uniformly recurrent}.
  Let $b,c \in \Sigma$ be two letters such that $b$ is \kl(morphic){growing}
  and $c$ appears infinitely often in $w$.
  Then, $c$ appears in every large enough infix of $w_{\geq N_0}$.
  Because $b$ appears infinitely often in $w$, 
  $f^n(b)$ is an infix of $w_{\geq N_0}$ for all $n$'s,
  and in particular $c$ appears in $f^n(c)$ for some $n$ large enough so that 
  $|f^n(b)|$ is larger than the bound of \kl{uniform recurrence} of $w_{\geq N_0}$.

  Let us now prove the second item. Let $u$ be a \kl(worphic){bounded word}
  that appears as an infix of $w$, by construction, it cannot contain
  any \kl(morphic){growing} letter, otherwise $|f^n(u)|$ would not be bounded.
  Because $w_{\geq N_0}$ is \kl{uniformly recurrent}, there exists a bound $M_0$ 
  on the distance between two consecutive \kl(morphic){growing} letters in 
  $w_{\geq N_0}$. Now, because $f$ is \kl(morphic){growing},
  there exists a \kl(morphic){growing} letter $b \in \Sigma$ that 
  appears infinitely often in $w$. As a consequence, the 
  \kl(morphic){bounded words} that appear as infixes of $w$ have length 
  at most $N_0 + M_0$.

  Conversely, assume that the two items are satisfied. Let us prove that $w$ is
  \kl{ultimately uniformly recurrent}. First, there exists $N_0$ such that
  every letter that appears in $w_{\geq N_0}$ appears infinitely often in
  $w_{\geq N_0}$ because $\Sigma$ is finite.

  \todo[inline]{finish the proof}
  following~\cite[Theorem 4]{NIPR09} we argue that $w_{\geq N_0}$ is
  \kl{uniformly recurrent}....

\begin{lemma}
  \label{morphic-uur:lem}
    Given a \kl{pure morphic sequence} $w \in \Sigma^{\Nat}$, one can decide
    whether it is \kl{ultimately uniformly recurrent}.
\end{lemma}
\begin{proof}
  Let us first assume that the morphism $f$ that generates $w$ is
  \kl(morphic){growing}. If it is not the case, then $w$ is 
  ultimately periodic, and a fortiori \kl{ultimately uniformly recurrent}.
  It is possible to decide whether $f$ is \kl(morphic){growing}
  by standard techniques on morphic words.

  On the one hand, one can guess a bound $N_0$ and check that 
  the suffix of the word $w_{\geq N_0}$ is \kl{uniformly recurrent},
  since the suffix of a \kl{morphic sequence} is also a \kl{morphic sequence},
  and \kl{uniform recurrence} is decidable for \kl{morphic sequences}
  \cite[Theorem 2]{DURAND13}.

  On the other hand, one can leverage the characterization of
  \cref{pure-morphic-uur:lem}. Indeed, it is shown in \cite[Proposition
  5]{NIPR09} that if the second item of \cref{pure-morphic-uur:lem} is not
  satisfied, then there exists a finite witness of that fact that can be
  computed from the morphism $f$. Similarly, to witness that the first item of
  \cref{pure-morphic-uur:lem} is not satisfied, one can guess two letters $a,b
  \in \Sigma$ and check that $a$ is \kl(morphic){growing}, both $a$ and $b$
  appear infinitely often in $w$, and that $b$ does not appear in $f^n(a)$ for
  any $n \in \Nat$. The first two conditions can be checked effectively, and
  the last one can be checked by noticing that if $b$ appears in $f^n(a)$ for
  some $n \in \Nat$, then it also appears in $f^m(a)$ for some $m \leq
  |\Sigma|$.
